---
output: github_document
---

```{r, echo = FALSE}
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
```

[![R-CMD-check](https://github.com/tsmodels/tsarma/workflows/R-CMD-check/badge.svg)](https://github.com/tsmodels/tsarma/actions)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg?style=flat-square)](commits/master)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/tsarma)](https://cran.r-project.org/package=tsarma)


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tsarma

The **tsarma** package implements methods for the maximum likelihood estimation of 
ARMA(p,q)-X models for stationary time series, backed by autodiff (using TMB package).
It includes methods for inference, filtering, prediction, simulation and backtesting.

It makes use of the [tsmodels](https://github.com/tsmodels) framework, providing 
common methods with similar inputs and classes with similar outputs.


The package vignette is available in the new tsmodels site [here](https://www.nopredict.com/packages/tsarma.html).

